The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at ESO/VLT under programme 088.A-0051 and 091.D-0904, at TNG under programme A24TAC\_38, at Gemini North under programme GN-2011B-Q-10 and GTC under programme GTC43-11B.}

### 2018-08-28 GRB111117A_corr.tex
Published

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

% Manually specified definitions
\input{definitions.tex}

@@

\input{authorlist.tex}

@@

\abstract{
It is notoriously difficult to localize short $\gamma$-ray bursts (sGRBs) and
their hosts to measure their redshifts. These measurements, however, are
critical to constrain the nature of sGRB progenitors, their redshift
distribution and the $r$-process element enrichment history of the universe.
Here, we present spectroscopy of the host galaxy of GRB~111117A and measure its
redshift to be $z = 2.211$. This makes GRB~111117A the most distant
high-confidence short duration GRB detected to date. Our spectroscopic redshift
supersedes a lower, previously estimated photometric redshift value for this
burst.

We use the spectroscopic redshift, as well as new imaging data to constrain the
nature of the host galaxy and the physical parameters of the GRB. The rest-frame
X-ray derived hydrogen column density, for example, is the highest compared to a
complete sample of sGRBs and seems to follow the evolution with redshift as
traced by the hosts of long GRBs.

The host lies in the brighter end of the expected sGRB host brightness
distribution at $z = 2.211$, and is actively forming stars. Using the observed
sGRB host luminosity distribution, we find that between 43 and 71 per cent of
all \textit{Swift}-detected sGRBs have too faint hosts at $z\sim2$ to allow for
a secure redshift determination. This implies that the measured sGRB redshift
distribution could be incomplete at high redshift. The large $z$ of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts (sGRBs) come
from the merger of a neutron star (NS), either with another NS, or a black hole,
due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Berger2013b, Yang2015, Jin2016, Rosswog2016}. This association was
recently confirmed by the simultaneous and co-spatial detection of gravitational
waves (GWs) from a binary neutron star merger and a sGRB
\citep{LIGOScientificCollaboration2017a, Goldstein2017, Savchenko2017}. The
absence of associated supernovae in deep searches
\citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b, Kann2011} supports this idea and
distinguishes the physical origin of sGRBs from their long-duration
counterparts, \citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}.

@@

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}. Unlike lGRBs,
sGRB afterglows are faint making absorption spectroscopy often ineffective.
Therefore, most of these measurements come from the associated hosts and is
potentially biased towards lower redshifts. Additionally, because the
\textit{Swift} sensitivity to sGRB decreases sharply with redshift
\citep{Behroozi2014}, the intrinsic redshift distribution of sGRBs is largely
unknown at higher redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above the detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long lived in comparison to lGRBs, with the host stellar mass affecting
the sGRB rates more than the star-formation rate (SFR) \citep{Berger2014}.

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers \citep{Ghirlanda2016}. This epochal breakthrough occurred
recently when the first NS-NS GW event ever was detected by LIGO/Virgo (GW
170817) and associated to the weak sGRB 170817A detected by the \textit{Fermi}
and INTEGRAL satellites \citep{LIGOScientificCollaboration2017a, Goldstein2017,
	Savchenko2017}. The simultaneous detection of a sGRB and a GW provides new
promising ways to constrain the binary inclination angle \citep{Arun2014,
	LIGOScientificCollaboration2017a} and measure cosmological distances
\citep{Nissanke2010, LIGOScientificCollaboration2017c}.

The total lifetime of NS binaries depends on their orbit, mass, spin, initial
separations and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated using host galaxy star-formation history models and
spatial distribution of sGRBs within their hosts \citep[][]{Berger2014}. The
most distant cosmological bursts, however, offer direct, hard limits on the
coalescence time scales.

We here present a spectrum of the host galaxy of the short GRB~111117A
($T_{90}=0.46$~s) and measure its redshift to be $z=2.211$. This value is
significantly higher than the previously estimated redshift based on photometric
studies \citep{Margutti2012,Sakamoto2013}. We present the GRB's rest frame
properties based on this new distance compared to previous analyses and revisit
the host properties derived from the new solution to the spectral energy
distribution (SED) fit. While no optical afterglow was detected, the excellent
localization from a detection of the X-ray afterglow by the \emph{Chandra} X-ray
Observatory allows us to discuss the positioning and environmental properties of
this remarkably distant sGRB. We use the $\Lambda$CDM cosmology parameters
provided by \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$. All magnitudes
are given in the AB system.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{
	FORS2 $R$-band imaging of the field of GRB~111117A with the X-shooter slit
overlaid. The slit position represents 4 epochs of spectroscopic observations
taken at similar position angles. The corresponding photometry is shown in
Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position as derived
from the \emph{Chandra} observations in \citet{Sakamoto2013}.
	}
	\label{fig:spec_setup}
\end{figure}

%\begin{figure}
%	\centering
%	\includegraphics[width=9cm]{figures/OIII_img.pdf}
%	\caption{2D-image of the \oiii$\lambda$5007 emission line. The location of a bright skyline is marked by the blue box. The location of the emission line is indicated with the red ellipse. Because the host is observed in nodding-mode, negative images of the emission line appear on both sides in the spatial direction.}
%	\label{fig:line}
%\end{figure}

\begin{figure*}
	\centering
	\includegraphics[width=18cm]{figures/grid.pdf}
	\caption{2D-images of the emission lines corresponding to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959, \oiii$\lambda$5007, and \ha. The location of bright skylines are marked by blue boxes. The locations of the emission lines are indicated with red ellipses. Because the host is observed in nodding-mode, negative images of the emission lines appear on both sides in the spatial direction. For the upper, left panel containing \lya, the systemic redshift position of \lya{} is marked by a yellow, vertical, dashed line. The red part of the \oii$\lambda$3727-doublet is affected by atmospheric absorption.}
	\label{fig:line}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/SEDspecphot.pdf}
	\caption{Best-fit SED to the derived photometry. The detection of \lya{}  is predicted from the SED fit and confirmed by the spectroscopic observations. Overplotted in grey is the observed spectrum, binned by 6 \AA{} for presentation purposes. Slit losses has been corrected for based on the average seeing of the observations, as confirmed by the comparison with the photometry. The blue, dashed line is the corresponding error spectrum, smoothed for presentation purposes. The reason for the spectral gaps at 5500 \AA{} and 10000 \AA{} is from the merging of the arms.}
	\label{fig:SED}
\end{figure*}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). Observations use a simple ABBA nodding pattern, with 5
\arcsec~nod throws. X-shooter covers the wavelength range from 3000~\AA{} to
24\,800~\AA{} (21\,000~\AA{} when the $K$-band blocking filter is used) across
three spectroscopic arms. We carried out the bias-correction, flat-fielding,
order tracing, wavelength calibration, rectification, and flux calibration using
the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} run
in physical mode. Because the echelle orders are curved across each detector, a
rectification algorithm is employed which introduces correlations between
neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the
UVB/VIS/NIR arm to minimize the degree of correlation while conserving the
maximal resolution. The observations are combined and extracted using scripts
described in Selsing et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}},
where the full spectral point spread function is modeled across each arm and
used for the optimal extraction algorithm \citep{Horne1986}. An overview of the
spectroscopic observations is given in Table~\ref{tab:spec_overview}, and the
slit position is shown in Fig.~\ref{fig:spec_setup}. Each of the epochs are
extracted individually and combined in a weighted fashion where the weight at
each pixel is chosen as median variance spectrum of the region surrounding that
pixel, thus avoiding to base the weight on the pixel variance. Slit-loss
correction has been applied on the combined spectrum based on the average seeing
of the observations. We show the extracted spectrum in Fig.~\ref{fig:SED}.

\input{tables/spec_overview.tex}

We determine a redshift of $z = 2.211 \pm 0.001$ from the simultaneous detection
of emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959,
\oiii$\lambda$5007, and \ha. \oii$\lambda$3727, \hb{}, and \oiii$\lambda$4959
are detected at low significance ($\sim$ 3-$\sigma$). The uncertainty on the
redshift is the standard deviation of independent measurements of the redshift
based on the individual line centroids (excluding \lya). We show cutouts of the
2D spectrum at the position of all the detected lines in Fig.~\ref{fig:line}. 
\ha{} is only visible in the first epoch due to the $K$-band blocking filter
used for the remaining observations. The nebular lines exhibit a spatial extent
of $\sim$ 1\farc5 and show significant velocity structure along the slit. A drop
in the continuum bluewards of the \lya{} line further supports the redshift. No
spectral evolution is observed across the epochs indicating that there is
negligible GRB afterglow contribution to the first epoch spectrum.

Using the luminosity of \ha, we can infer the SFR of the host
\citep{Kennicutt1998}. At the redshift of the GRB host, \ha{} is observed at
around 21\,000~\AA{} where the night sky is very bright. In addition, several
bright sky-lines are superposed on the line, making an accurate estimate of the
\ha~flux difficult. Due to their velocity structure, the lines exhibit clear
deviations from a Gaussian and given the low S/N of the spectra we do not
attempt any parametric fits. We instead obtain a limit on the SFR by numerically
integrating the part of \ha{} free of contamination and obtain $F_{\mathrm{H,X} \alpha} >
4.1 \times 10^{-17}~\mathrm{erg}~\mathrm{s}^{-1}~\mathrm{cm}^{-2}$. After
converting the \citet{Kennicutt1998} relation to a \citet{Chabrier2003} initial
mass function using the conversion factor from \citet{Madau2014}, we derive a
limit of $\mathrm{SFR} > 7 M_\odot$~yr$^{-1}$. We additionally obtain the \lya{} line
flux by numerically integrating the entire \lya{} line complex and obtain
$F_{\mathrm{Ly}\alpha} = (2.0 \pm 0.5) \times 10^{-17}~\mathrm{erg}~
\mathrm{s}^{-1}~\mathrm{cm}^{-2}$, where the error on the flux is found by
integrating the associated error spectrum over the same spectral region.

@@

From the SED-fit (Sect.~\ref{SED}) the host is constrained to contain very
little or no dust. Consistently, \lya{} is detected although its presence does
not exclude dust. Therefore we do not apply a dust-correction to the measured
\ha{} flux here. \oii{} is close to a region of strong telluric absorption,
which is why no SFR is inferred from this line.

The total extent of the lines in velocity space is $\sim$ 450 km s$^{-1}$. The line
profiles shows an asymmetric "double-horned" profile, indicating that we are
seeing a galaxy with a large degree of coherent rotational motion relative to
the line-of-sight. If we assume that we are viewing a spiral galaxy edge-on,
this is a measure of the rotational velocity of the gas. If we assume that the
spectral resolution and the turbulent width of the lines are negligible compared
to the rotational velocity, we can, based on the projected size of the source
and the width of the lines, put a constraint on the dynamical mass of the galaxy
\citep{DeBlok2014}. Based on the physical size along the slit and the velocity
width of \oiii$\lambda5007$, we infer $M_\text{dyn} \gtrsim 10^{10.8}$
$M_\odot$. Because we are viewing the host inclined at an angle relative to
edge-on and because the slit is not aligned along the long axis of the host,
this value is a lower limit.

@@

In addition to the spectroscopy presented above, we imaged the field of
GRB~111117A in multiple broad-band filters using the VLT equipped with FORS2
($gRIz$ filters) and HAWK-I ($JHK_{\mathrm{s}}$ filters). These new data are
complemented by a re-analysis of some of the imaging used in
\citet{Margutti2012} and \citet{Sakamoto2013} that are available to us (GTC
$gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in
Table~\ref{tab:phot_overview}. Most data were taken long after the GRB had faded
when no afterglow contribution was present. Given the faintness of the afterglow
(see Sect. \ref{xray}, and \citealt{Cucchiara2011, Cenko2011}), we also expect
negligible contribution to the earliest epochs, which is confirmed by the
consistency between the two $g$-band measurements.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[longauth]{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

% Manually specified definitions
\newcommand{\farc}{\hbox{$.\!\!^{\prime\prime}$}} 
\newcommand{\ergA}{$\rm{erg\,cm^{-2}\,s^{-1}\,\AA^{-1}}$} 
\newcommand{\erg}{$\rm{erg\,cm^{-2}\,s^{-1}}$} 
\newcommand{\kms}{$\rm{km\,s^{-1}}\,$}

\newcommand{\griz}{$g' r' i' z'$}
\newcommand{\JHK}{$JHK_{\rm{s}}$}
\newcommand{\gK}{$g' r' i' z' JHK_{\rm{s}}$}
\newcommand{\Msun}{$M_\odot$}

% Elements
\newcommand{\lya}{Ly$\alpha$} 
\newcommand{\lyb}{Ly$\beta$} 
\newcommand{\lyg}{Ly$\gamma$} 
\newcommand{\hb}{H$\beta$} 
\newcommand{\ha}{H$\alpha$} 
\newcommand{\hg}{H$\gamma$} 
\newcommand{\hi}{\ion{H}{i}} 
\newcommand{\hei}{[\ion{He}{i}]} 
\newcommand{\oi}{[\ion{O}{i}]} 
\newcommand{\sii}{[\ion{S}{ii}]} 
\newcommand{\siii}{[\ion{S}{iii}]} 
\newcommand{\oii}{[\ion{O}{ii}]} 
\newcommand{\oiii}{[\ion{O}{iii}]}
\newcommand{\nii}{[\ion{N}{ii}]} 
\newcommand{\nv}{[\ion{N}{v}]} 
\newcommand{\neiii}{[\ion{Ne}{iii}]} 
\newcommand{\NIii}{[\ion{Ni}{ii}]} 
\newcommand{\feii}{\ion{Fe}{ii}} 
\newcommand{\civ}{\ion{C}{iv}} 
\newcommand{\cii}{\ion{C}{ii}}
\newcommand{\mgi}{\ion{Mg}{i}} 
\newcommand{\mgii}{\ion{Mg}{ii}} 
\newcommand{\ali}{\ion{Al}{i}} 
\newcommand{\alii}{\ion{Al}{ii}} 
\newcommand{\aliii}{\ion{Al}{iii}} 
\newcommand{\SIi}{\ion{Si}{i}} 
\newcommand{\SIii}{\ion{Si}{ii}} 
\newcommand{\SIiii}{\ion{Si}{iii}} 
\newcommand{\SIiv}{\ion{Si}{iv}} 
\newcommand{\znii}{\ion{Zn}{ii}} 
\newcommand{\crii}{\ion{Cr}{ii}} 
\newcommand{\mnii}{\ion{Mn}{ii}} 
\newcommand{\tiii}{\ion{Ti}{ii}} 
\newcommand{\caii}{\ion{Ca}{ii}}

@@

\author{J.~Selsing\inst{1}
        \and T.~Kr\"{u}hler\inst{2}
        \and D.~Malesani\inst{1}
        \and P.~D'Avanzo\inst{3}
        \and S.~Schulze \inst{4, 5, 6}
        \and S.~D.~Vergani\inst{3, 7}
        \and J.~Palmerio\inst{8}
        \and J.~Japelj\inst{9}
        \and B.~Milvang-Jensen\inst{1}
        \and D.~Watson\inst{1}
        \and P.~Jakobsson\inst{10}
        %
        \and J.~Bolmer \inst{2}
        \and Z.~Cano\inst{11}
        \and S.~Covino \inst{3}
        %\and L.~B.~Christensen\inst{1}
        \and V.~D'Elia\inst{12, 13}
        \and A.~de~Ugarte~Postigo\inst{11, 1}
        \and J.~P.~U.~Fynbo\inst{1}
        \and A.~Gomboc\inst{14}
        \and K.~E.~Heintz\inst{10,1}
        \and L.~Kaper \inst{9}
        \and A.~J.~Levan \inst{15}
        \and S.~Piranomonte \inst{12}
        \and G.~Pugliese \inst{9}
        \and R.~S\'{a}nchez-Ram\'{\i}rez \inst{11, 16}
        \and M.~Sparre\inst{1,17}
        \and N.~R.~Tanvir\inst{18}
        \and C.~C.~Th\"{o}ne\inst{11}
        \and K.~Wiersema \inst{18}
}

\institute{Dark Cosmology Centre, Niels Bohr Institute, University of Copenhagen, Juliane Maries Vej 30, 2100 K\o benhavn \O, Denmark
        %2
        \and Max-Planck-Institut f\"{u}r extraterrestrische Physik, Giessenbachstra\ss e, 85748 Garching, Germany
        %3
        \and INAF - Osservatorio Astronomico di Brera, via E. Bianchi 46, I-23807, Merate (LC), Italy
        %4
        \and Instituto de Astrof\'isica, Facultad de F\'isica, Pontificia Universidad Cat\'olica de Chile, Vicu\~{n}a Mackenna 4860, 7820436 Macul, Santiago, Chile
        %5
        \and Millennium Institute of Astrophysics, Vicu\~{n}a Mackenna 4860, 7820436 Macul, Santiago, Chile
        %6
        \and Department of Particle Physics and Astrophysics, Faculty of Physics, Weizmann Institute of Science, Rehovot 76100, Israel
        %7
        \and GEPI, Observatoire de Paris, PSL Research University, CNRS, Place Jules Janssen, 92190 Meudon, France
        %8
        \and Sorbonne Universités, UPMC Univ. Paris 6 et CNRS, UMR 7095, Institut d’Astrophysique de Paris, 98 bis bd Arago, 75014 Paris, France
        %9
        \and Anton Pannekoek Institute for Astronomy, University of Amsterdam, Science Park 904, 1098 XH Amsterdam, The Netherlands
        %10
        \and Centre for Astrophysics and Cosmology, Science Institute, University of Iceland, Dunhagi 5, 107 Reykjav\'ik, Iceland
        %11
        \and Instituto de Astrof\'isica de Andaluc\'ia (IAA-CSIC), Glorieta de la Astronom\'ia s/n, E-18008, Granada, Spain
        %12
        \and INAF-Osservatorio Astronomico di Roma, Via Frascati 33, I-00040 Monteporzio Catone, Italy
        %13
        \and ASI-Science Data Centre, Via del Politecnico snc, I-00133 Rome, Italy
        %14
        \and Centre for Astrophysics and Cosmology, University of Nova Gorica, Vipavska 13, 5000 Nova Gorica, Slovenia.
        %15
        \and Department of Physics, University of Warwick, Coventry CV4 7AL, UK
        %16
        \and Istituto de Astrofisica e Planetologia Spaziali, INAF, Via Fosso del Cavaliere 100, I-00133 Roma, Italy
        %17
        \and Heidelberger Institut f{\"u}r Theoretische Studien, Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany 
        %18
        \and Department of Physics \& Astronomy and Leicester Institute of Space \& Earth Observation,
        University of Leicester, University Road, Leicester, LE1 7RH, UK
}

@@

\abstract{
It is notoriously difficult to localize short $\gamma$-ray bursts (sGRBs) and
their hosts to measure their redshifts. These measurements, however, are
critical for  constraining the nature of sGRB progenitors, their redshift
distribution, and the $r$-process element enrichment history of the universe.
Here we present spectroscopy of the host galaxy of GRB~111117A and measure its
redshift to be $z = 2.211$. This makes GRB~111117A the most distant
high-confidence short duration GRB detected to date. Our spectroscopic redshift
supersedes a lower, previously estimated photometric redshift value for this
burst.

We use the spectroscopic redshift, as well as new imaging data to constrain the
nature of the host galaxy and the physical parameters of the GRB. The rest-frame
X-ray derived hydrogen column density, for example, is the highest compared to a
complete sample of sGRBs and seems to follow the evolution with redshift as
traced by the hosts of long GRBs. From the detection of \lya~emission in the
spectrum, we are able to constrain the escape fraction of \lya~in the host.

The host lies in the brighter end of the expected sGRB host brightness
distribution at $z = 2.211$, and is actively forming stars. Using the observed
sGRB host luminosity distribution, we find that between 43\%\ and 71\%\ of
all \textit{Swift}-detected sGRBs have hosts that are  too faint  at $z\sim2$ to allow for
a secure redshift determination. This implies that the measured sGRB redshift
distribution could be incomplete at high redshift. The high $z$ of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts (sGRBs) come
from the merger of a neutron star (NS)  either with another NS or with a black hole,
due to their apparent association with kilonovae \citep{Barnes2013a,
        Tanvir2013b, Berger2013b, Yang2015, Jin2016, Rosswog2016}. This association has
recently been confirmed by the simultaneous and co-spatial detection of gravitational
waves (GWs) from a binary neutron star merger and a sGRB
\citep{LIGOScientificCollaboration2017a, Goldstein2017, Savchenko2017}; however,
to what degree the sGRB associated with GW170817 is a cosmological sGRB remains
an open question \citep{Lyman2018, Margutti2018}. The absence of associated
supernovae in deep searches \citep[e.g.,][]{Hjorth2005a,Fox2005,Hjorth2005b,
        Kann2011} supports this idea and distinguishes the physical origin of sGRBs from
their long-duration counterparts \citep[but see also][]{Fynbo2006b,
        Valle2006, Gal-Yam2006}.

@@

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}. Unlike lGRBs,
sGRB afterglows are faint, making absorption spectroscopy often ineffective.
Therefore, most of these measurements come from the associated hosts and is
potentially biased towards lower redshifts. Additionally, because the
\textit{Swift} sensitivity to sGRB decreases sharply with redshift
\citep{Behroozi2014}, the intrinsic redshift distribution of sGRBs is largely
unknown at higher redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
 star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases
no host galaxy can be identified above the detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long-lived in comparison to lGRBs, with the host stellar mass affecting
the sGRB rates more than the star formation rate (SFR) \citep{Berger2014}.

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers \citep{Ghirlanda2016}. This epochal breakthrough occurred
recently when the first ever NS--NS GW event  was detected by LIGO/Virgo (GW
170817) and associated with the weak sGRB 170817A detected by the \textit{Fermi}
and INTEGRAL satellites \citep{LIGOScientificCollaboration2017a, Goldstein2017,
        Savchenko2017}. The simultaneous detection of a sGRB and a GW provides new
promising ways to constrain the binary inclination angle \citep{Arun2014,
        LIGOScientificCollaboration2017a} and to measure cosmological distances
\citep{Nissanke2010, LIGOScientificCollaboration2017c}.

The total lifetime of NS binaries depends on their orbit, mass, spin, initial
separations, and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated using host galaxy star formation history models and
spatial distribution of sGRBs within their hosts \citep[][]{Berger2014}. The
most distant cosmological bursts, however, offer direct, hard limits on the
coalescence timescales.

Here we  present a spectrum of the host galaxy of the short GRB~111117A
($T_{90}=0.46$~s) and measure its redshift to be $z=2.211$. This value is
significantly higher than the previously estimated redshift of $z \sim 1.3$ based on photometric
studies \citep{Margutti2012,Sakamoto2013}. We present the GRB's rest frame
properties based on this new distance compared to previous analyses and revisit
the host properties derived from the new solution to the spectral energy
distribution (SED) fit. While no optical afterglow was detected, the excellent
localization from a detection of the X-ray afterglow by the \emph{Chandra} X-ray
Observatory allows us to discuss the positioning and environmental properties of
this remarkably distant sGRB. We use the $\Lambda$CDM cosmology parameters
provided by \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$. All magnitudes
are given in the AB system.

All data, code, and calculations related to the paper along with the
paper itself are available \href{https://github.com/jselsing/GRB111117A}{online}.

@@

\begin{figure}
        \centering
       \includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
        \caption{
        FORS2 $R$-band imaging of the field of GRB~111117A with the X-shooter slit
overlaid. The slit position represents four epochs of spectroscopic observations
taken at similar position angles. The corresponding photometry is shown in
Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position (R.A., Dec.) (J2000) = (00$^{\rm h}$ 50$^{\rm m}$ 46.264$^{\rm s}$, +23$^{\circ}$ 00$^{\prime}$ 39.98$^{\prime\prime}$) as derived
from the \emph{Chandra} observations in \citet{Sakamoto2013}.
        }
        \label{fig:spec_setup}
\end{figure}

%\begin{figure}
%       \centering
%%      \includegraphics[width=9cm]{figures/OIII_img.pdf}
%       \caption{2D-image of the \oiii$\lambda$5007 emission line. The location of a bright skyline is marked by the blue box. The location of the emission line is indicated with the red ellipse. Because the host is observed in nodding-mode, negative images of the emission line appear on both sides in the spatial direction.}
%       \label{fig:line}
%\end{figure}

\begin{figure*}
        \centering
        \includegraphics[width=18cm]{figures/grid.pdf}
        \caption{Two-dimensional  images of the emission lines corresponding to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959, \oiii$\lambda$5007, and \ha. The location of bright skylines are marked by blue boxes. The locations of the emission lines are indicated with red ellipses. Because the host is observed in nodding mode, negative images of the emission lines appear on both sides in the spatial direction. For the upper left panel containing \lya, the systemic redshift position of \lya{} is marked by a yellow vertical dashed line. The red part of the \oii$\lambda$3727-doublet is affected by atmospheric absorption.}
        \label{fig:line}
\end{figure*}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO program 088.A-0051 (PI: Fynbo) and again later under ESO program
091.D-0904 (PI: Hjorth). The observations used a simple ABBA nodding pattern, with 5
\arcsec~nod throws. The X-shooter covers the wavelength range from 3000~\AA{} to
24\,800~\AA{} (21\,000~\AA{} when the $K$-band blocking filter is used) across
three spectroscopic arms. We carried out the bias-correction, flat-fielding,
order tracing, wavelength calibration, rectification, and flux calibration using
the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} run
in physical mode. Because the echelle orders are curved across each detector, a
rectification algorithm is employed that introduces correlations between
neighboring pixels. We selected a pixel scale of 0.2/0.2/0.6 \AA/pix for the
UVB/VIS/NIR arm to minimize the degree of correlation while conserving the
maximum resolution. The observations were combined and extracted using the scripts
described in \citet{2018arXiv180207727S} and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}},
where the full spectral point spread function was modeled across each arm and
used for the optimal extraction algorithm \citep{Horne1986}. An overview of the
spectroscopic observations is given in Table~\ref{tab:spec_overview}, and the
slit position is shown in Fig.~\ref{fig:spec_setup}. Each of the epochs was
extracted individually and combined in a weighted fashion where the weight at
each pixel was chosen as median variance spectrum of the region surrounding that
pixel, thus avoiding basing the weight on the pixel variance. Slit-loss
correction was applied on the combined spectrum based on the average seeing
of the observations. We show the extracted spectrum in Fig.~\ref{fig:SED}.

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/SEDspecphot.pdf}
	\caption{Best-fit SED to the derived photometry. The detection of \lya{}  is predicted from the SED fit and confirmed by the spectroscopic observations. Overplotted in gray is the observed spectrum, binned by 6 \AA{} for presentation purposes. Slit losses have been corrected for based on the average seeing of the observations, as confirmed by the comparison with the photometry. The blue dashed curve is the corresponding error spectrum, smoothed for presentation purposes. The spectral gaps at 5500 \AA{} and 10000 \AA{} are due to the merging of the arms.}
	\label{fig:SED}
\end{figure*}

We  determine a redshift of $z = 2.211 \pm 0.001$ from the simultaneous detection
of emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959,
\oiii$\lambda$5007, and \ha. The \oii$\lambda$3727, \hb{}, and \oiii$\lambda$4959
lines are detected at low significance ($\sim$ 3-$\sigma$). The uncertainty on the
redshift is the standard deviation of independent measurements of the redshift
based on the individual line centroids (excluding \lya). We show cutouts of the
2D spectrum at the position of all the detected lines in Fig.~\ref{fig:line}. The 
\ha{} line is only visible in the first epoch due to the $K$-band blocking filter
used for the remaining observations. The nebular lines exhibit a spatial extent
of $\sim$ 1\farc5 and show significant velocity structure along the slit. A drop
in the continuum bluewards of the \lya{} line further supports the redshift. No
spectral evolution is observed across the epochs indicating that there is
negligible GRB afterglow contribution to the first epoch spectrum.

Using the luminosity of \ha, we can infer the SFR of the host
\citep{Kennicutt1998}. At the redshift of the GRB host, \ha{} is observed at
around 21\,000~\AA{} where the night sky is very bright. In addition, several
bright sky-lines are superposed on the line, making it difficult to obtain  an accurate estimate of the
\ha~flux. Due to their velocity structure, the lines exhibit clear
deviations from a Gaussian and given the low S/N of the spectra we do not
attempt any parametric fits. We instead obtain a limit on the SFR by numerically
integrating the part of \ha{} free of contamination and obtain $F_{\mathrm{H,X} \alpha} >
4.1 \times 10^{-17}~\mathrm{erg}~\mathrm{s}^{-1}~\mathrm{cm}^{-2}$. After
converting the \citet{Kennicutt1998} relation to a \citet{Chabrier2003} initial
mass function using the conversion factor from \citet{Madau2014}, we derive a
limit of $\mathrm{SFR} > 7 M_\odot$~yr$^{-1}$. We additionally obtain the \lya{} line
flux by numerically integrating the entire \lya{} line complex and obtain
$F_{\mathrm{Ly}\alpha} = (2.0 \pm 0.5) \times 10^{-17}~\mathrm{erg}~
\mathrm{s}^{-1}~\mathrm{cm}^{-2}$, where the error on the flux is found by
integrating the associated error spectrum over the same spectral region.

@@

From the SED fit (Sect.~\ref{SED}) the host is constrained to contain very
little or no dust. Consistently, \lya{} is detected although its presence does
not exclude dust. Therefore, we do not apply a dust-correction to the measured
\ha{} flux here. The  \oii{} line is close to a region of strong telluric absorption,
which is why no SFR is inferred from it.

The total extent of the lines in velocity space is $\sim$ 450 km s$^{-1}$. The line
profiles shows an asymmetric ``double-horned'' profile, indicating that we are
seeing a galaxy with a large degree of coherent rotational motion relative to
the line of sight. If we assume that we are viewing a spiral galaxy edge-on,
this is a measure of the rotational velocity of the gas. If we assume that the
spectral resolution and the turbulent width of the lines are negligible compared
to the rotational velocity, based on the projected size of the source
and the width of the lines we can put a constraint on the dynamical mass of the galaxy
\citep{DeBlok2014}. Based on the physical size along the slit and the velocity
width of \oiii$\lambda5007$, we infer $M_\text{dyn} \gtrsim 10^{10.8}$
$M_\odot$. Because we are viewing the host inclined at an angle relative to
edge-on and because the slit is not aligned along the long axis of the host,
this value is a lower limit.

@@

In addition to the spectroscopy presented above, we imaged the field of
GRB~111117A in multiple broadband filters using the VLT equipped with FORS2
($gRIz$ filters) and HAWK-I ($JHK_{\mathrm{s}}$ filters). These new data are
complemented by a re-analysis of some of the imaging used in
\citet{Margutti2012} and \citet{Sakamoto2013} that are available to us (GTC
$gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in
Table~\ref{tab:phot_overview}. Most of the data were taken long after the GRB had faded
when no afterglow contribution was present. Given the faintness of the afterglow
(see Sect. \ref{xray}, and \citealt{Cucchiara2011, Cenko2011}), we also expect
negligible contribution to the earliest epochs, which is confirmed by the
consistency between the two $g$-band measurements.
